---
title: "Kobe Bryant Hot Hand Analysis"
author: "Akeem Lawrence"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(openintro)

set.seed(1986)
```

# Overview

Basketball players who make several consecutive shots are often described as having a **hot hand**. The hot hand hypothesis suggests that making a shot increases the probability of making the next one, meaning successive shot outcomes may not be independent.

This analysis examines Kobe Bryant's shooting performance during the 2009 NBA Finals and compares his observed shooting streaks with those generated by a simulated independent shooter.

The central question is:

> Do Kobe Bryant's shooting streaks provide evidence of a hot hand, or are they consistent with what could occur when each shot is independent?

# Data

The analysis uses the `kobe_basket` dataset from the `openintro` package.

```{r inspect-data}
glimpse(kobe_basket)
```

The dataset contains 133 Kobe Bryant shot attempts from the 2009 NBA Finals.

Each observation includes information such as:

- Opponent
- Game
- Quarter
- Game time
- Shot description
- Shot result

The `shot` variable identifies each attempt as:

- `H` = Hit
- `M` = Miss

```{r shot-results}
table(kobe_basket$shot)
```

# Defining a Shooting Streak

For this analysis, a shooting streak is defined as the number of consecutive made baskets before a miss occurs.

For example:

`H M | M | H H M | M | M | M`

This sequence contains streak lengths of:

`1, 0, 2, 0, 0, 0`

A streak length of `0` represents a miss that was not preceded by a made basket within that streak.

# Kobe Bryant's Shooting Streaks

The `calc_streak()` function is used to calculate Kobe Bryant's observed streak lengths.

```{r kobe-streak}
kobe_streak <- calc_streak(kobe_basket$shot)

summary(kobe_streak$length)
max(kobe_streak$length)
```

```{r plot-kobe-streak}
ggplot(kobe_streak, aes(x = length)) +
  geom_bar() +
  labs(
    title = "Kobe Bryant Shooting Streak Distribution",
    subtitle = "2009 NBA Finals",
    x = "Consecutive Made Baskets Before a Miss",
    y = "Frequency"
  ) +
  theme_minimal()
```

Kobe's most common streak length was `0`, and his longest observed streak contained **4 consecutive made baskets**.

The presence of several shooting streaks may appear consistent with the hot hand idea, but streaks alone do not demonstrate dependence between shots. Similar streaks can occur naturally through random variation.

# Independent Shooter Simulation

To create a comparison baseline, I simulated an independent shooter taking the same number of shots as Kobe.

The simulated shooter has:

- 133 shot attempts
- 45% probability of making each shot
- 55% probability of missing each shot
- Independent shot outcomes

```{r simulate-shooter}
shot_outcomes <- c("H", "M")

sim_basket <- sample(
  shot_outcomes,
  size = 133,
  replace = TRUE,
  prob = c(0.45, 0.55)
)

table(sim_basket)
```

Because each simulated attempt is independent, the outcome of one shot has no effect on the probability of making the next shot.

# Simulated Shooting Streaks

```{r simulated-streak}
sim_streak <- calc_streak(sim_basket)

summary(sim_streak$length)
max(sim_streak$length)
```

```{r plot-simulated-streak}
ggplot(sim_streak, aes(x = length)) +
  geom_bar() +
  labs(
    title = "Independent Shooter Streak Distribution",
    subtitle = "133 simulated shots at 45% shooting",
    x = "Consecutive Made Baskets Before a Miss",
    y = "Frequency"
  ) +
  theme_minimal()
```

The simulation shows that an independent shooter can also produce multiple consecutive made baskets.

This is important because it demonstrates that shooting streaks can occur even when the probability of making each shot remains constant.

# Comparing Kobe with the Independent Shooter

```{r compare-distributions}
ggplot() +
  geom_density(
    data = kobe_streak,
    aes(x = length, fill = "Kobe Bryant"),
    alpha = 0.5
  ) +
  geom_density(
    data = sim_streak,
    aes(x = length, fill = "Independent Shooter"),
    alpha = 0.5
  ) +
  labs(
    title = "Comparison of Shooting Streak Distributions",
    x = "Streak Length",
    y = "Density",
    fill = "Shooter"
  ) +
  theme_minimal()
```

Kobe's streak distribution and the independent shooter's simulated distribution both contain short shooting streaks as well as occasional longer sequences of made baskets.

At first glance, Kobe's consecutive makes might appear to support the hot hand hypothesis. However, the independent simulation demonstrates that similar streaks can arise without any increase in shooting probability after a made basket.

# Interpretation

The analysis does **not provide strong evidence that Kobe Bryant's observed shooting streaks require a hot hand explanation**.

Kobe recorded multiple streaks of consecutive made shots, including a longest streak of four makes. However, an independent shooter with a comparable shooting probability can also generate meaningful streaks simply through random variation.

Therefore, observing a shooting streak is not by itself sufficient evidence that making one shot increases the probability of making the next.

A stronger statistical investigation would require evaluating Kobe's observed streak patterns across many simulated sequences rather than relying on a single independent-shooter simulation.

# Limitations

This analysis has several limitations:

1. It compares Kobe's actual performance with only one simulated sequence.
2. The assumed shooting probability is fixed at 45%.
3. Shot difficulty, location, defensive pressure, game situation, and fatigue are not modeled.
4. Streak length alone does not directly measure the conditional probability of making a shot after a previous make.
5. The dataset contains only 133 shot attempts from one NBA Finals series.

# Future Analysis

A more rigorous extension could:

- Run thousands of independent-shooter simulations
- Compare Kobe's longest observed streak with the simulated distribution of longest streaks
- Estimate a simulation-based p-value
- Compare shooting percentage after a make versus after a miss
- Analyze streak behavior across multiple seasons
- Compare multiple NBA players
- Incorporate shot location and shot difficulty

# Conclusion

Kobe Bryant produced several notable shooting streaks during the 2009 NBA Finals, but this analysis shows that streaks of consecutive made baskets can also emerge from an independent shooting process.

The results therefore illustrate an important statistical lesson in basketball analytics:

> A visible streak does not necessarily imply that the probability of success has changed.

Simulation provides a useful benchmark for separating patterns that appear meaningful from patterns that may arise through random variation.
